### 2017-04-13 paper/oopsla17-submission/learning.tex
done with overview

% \begin{figure}[ht]
% \begin{minipage}{0.6\linewidth}
% \begin{code}
%   let rec sumList xs =
%     match xs with
%     | []     -> []
%     | hd::tl -> hd + sumList tl
% \end{code}
% \end{minipage}
% \begin{minipage}{0.3\linewidth}
% \begin{code}[numbers=left]
% []
% hd + sumList tl
% sumList tl
% tl
% \end{code}
% \end{minipage}
% \caption{An ill-typed program (left) and a selection of its
%   sub-expressions (right).}
% \label{fig:sumList}
% \end{figure}
\begin{table}[ht]
\caption{Example Feature Vectors}\label{tab:sumList}
\begin{tabular}{lrrrrrr}
\toprule
Expression
  & \IsNil & \IsCaseListP & \ExprSize
  & \HasTypeIntCOne & \HasTypeList & \InSlice \\
\midrule
|[]|
  & 1 & 1 & 1 & 0 & 1 & 1 \\
|hd + sumList tl|
  & 0 & 1 & 5 & 1 & 0 & 1 \\
|sumList tl|
  & 0 & 0 & 3 & 0 & 1 & 1 \\
|tl|
  & 0 & 0 & 1 & 0 & 1 & 0 \\
\bottomrule
\end{tabular}
\bigskip
\caption*{A selection of the features we would extract from the
\lstinline!sumList! program in \autoref{fig:sumList}. A feature is
considered \emph{enabled} if it has a non-zero value, and
\emph{disabled} otherwise. A ``-P'' suffix indicates that the feature
describes the parent of the current expression, a ``-C$n$'' suffix
indicates that the feature describes the $n$-th (left-to-right) child of
the current expression.  Note that, since we rely on a partial typing
derivation, we are subject to the well-known traversal bias and label
the expression \lstinline!sumList tl! as having type
$\tlist{\cdot}$. The model will have to learn to correct for this bias.}
\end{table}
The first issue we must tackle is formulating our learning task in
machine learning terms.
%
We are given programs over \lang, but the learning algorithms expect to work
with \emph{feature vectors} $\V$ --- vectors of real numbers, where each
column describes a particular aspect of the input.
%
Thus, our first task is to convert programs to feature vectors.

>>>>

% \begin{figure}[ht]
% \begin{minipage}{0.6\linewidth}
% \begin{code}
%   let rec sumList xs =
%     match xs with
%     | []     -> []
%     | hd::tl -> hd + sumList tl
% \end{code}
% \end{minipage}
% \begin{minipage}{0.3\linewidth}
% \begin{code}[numbers=left]
% []
% hd + sumList tl
% sumList tl
% tl
% \end{code}
% \end{minipage}
% \caption{An ill-typed program (left) and a selection of its
%   sub-expressions (right).}
% \label{fig:sumList}
% \end{figure}

\input{feature-table}

The first issue we must tackle is formulating our learning task in
machine learning terms.
%
We are given programs over \lang, but the learning algorithms expect to work
with \emph{feature vectors} $\V$ --- vectors of real numbers, where each
column describes a particular aspect of the input.
%
Thus, our first task is to convert programs to feature vectors.

### 2017-04-13 paper/oopsla17-submission/overview.tex
done with overview

The first step is to create a training
set of ill-typed programs, where each
erroneous sub-term is explicitly labeled.
%
Prior work has often enlisted
expert users to curate a set of
ill-typed programs and then
\emph{manually} determine the
correct fix~\citep[\eg][]{Lerner2007-dt,Loncaric2016-uk}.
%
This method is suitable for evaluating
the quality of a localization (or repair)
algorithm on a small number (\eg, 10-100s)
of programs.
%
However, in general it requires a great
deal of effort for the expert to divine
the original programmers' intentions.
%
Consequently, is difficult to scale the
expert-labeling to yield a dataset large
enough (\eg, 1000s of programs) needed to
facilitate machine learning.
%
More importantly, this approach fails to
capture the \emph{frequency} with which
errors occur in practice.

We solve both the scale and frequency problems
by instead extracting blame-labeled data sets
from \emph{interaction traces}.
%
Software development is an iterative process.
Programmers, perhaps after a lengthy (and
sometimes frustrating) back-and-forth with
the type checker, eventually end up fixing
their own programs.
%
Thus, we implemented an instrumented
version of the \ocaml compiler that records
this conversation, \ie records the sequence
of programs submitted by each programmer and
whether or not it was deemed type-correct.
%
For each ill-typed program in
a particular programmer's trace,
we find the \emph{first subsequent}
program in the trace that type checks
and declare it to be the fixed version.
%
From this pair of an ill-typed program
and its fix, we can extract a \emph{diff}
of the abstract syntax trees, and then assign
the blame labels to the \emph{smallest}
sub-tree in the diff.

For example, suppose our student
fixed the |sumList| program in
\autoref{fig:sumList} by replacing
|[]| with |0|, the diff would
include only the |[]| expression.
%
Thus we would determine that the
|[]| expression (and \emph{not} the
|+| or the recursive call |sumList t|)
is to blame.

@@

@@

@@

% Evaluation

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\paragraph{\textbf{Challenge (1): How to represent programs?}}

The first and prime challenge we address is how to translate programs
into a format favorable to learning a model of type errors.
%
Programs can be represented in various ways, from the raw text to
richly-structured abstract syntax trees and control-flow graphs, but
many representations are not directly amenable to machine learning.
%
Furthermore, program representations that are amenable to one learning
task, \eg code completion, may not be favorable to our problem of
assigning blame for type errors.

Machine learning techniques typically take as input \emph{feature
vectors}, fixed-length vectors of real numbers where each element
represents some property of the input.
%
Thus, to apply machine learning to our problem we must
determine:
%
(1) which features of a program would be useful for modeling type errors; and
%
(2) how can we represent a variable-size program as a fixed-length vector?

\paragraph{\textbf{Solution: Vectorized Abstract Syntax Trees}}
\ES{name up for grabs...}
We propose to extract features from the program's
abstract syntax tree (AST), a natural choice since the type checker
operates on ASTs.
%
We model features as \emph{queries} over the AST.\@
%
Our initial set of interesting features includes the syntactic class of
each expression.
%
That is, the classifier should be able to distinguish between the |[]|
and |+| expressions in \autoref{fig:sumList} because they represent
different syntactic expression forms.
%
Modeling the syntactic class of an expression gives the classifier a
basic notion of the relative frequency of blame assignment for the
various program elements, \ie perhaps |[]| is \emph{empirically} more
likely to be blamed than |+|.
% % ES: (2) doens't really make sense here
% (2) the type checker uses the syntactic class of an expression to
%     impose typing constraints on the neighboring expressions.

A programming language has a fixed set of AST elements,\footnote{We will
  discuss handling arbitrary, user-defined types in
  \autoref{sec:discussion}.} which suggests a natural way to obtain
fixed-size feature vectors. Rather than translating a program into a
single feature vector, we translate it into a \emph{set} of feature
vectors, one for each expression in the program.
%
Thus, the |[]| and |+| expressions will each be given their own feature
vector, and we will ask the classifier to predict for each
\emph{independently} whether it should be blamed.

Of course, each expression occurs in some \emph{context}, and we would
like the classifier to be able make decisions based on the context as
well.
%
The context is particularly important for our task as each expression
imposes typing constraints on its neighbors, \eg a |+| operator tells
the type checker that both children must have type |int| and that
the parent must accept an |int|.
%
We might expect that the recursive call |sumList t| is
unlikely to be at fault because |sumList| has a function type --- the
user may have called the wrong function or supplied the wrong argument,
but the application itself is probably correct.
%
Conversely, if the student wrote |h sumList t| (\ie forgot the |+|), we
might wish to blame the application rather than |h| because |h|
\emph{does not} have a function type.
%
Thus we extend each expression's feature vector with
\emph{contextual features} of its parent and children.

Our contextual features include the syntactic class of the neighboring
expressions, their inferred types (when available), and, crucially, whether
the expression occurs in a minimal type error \emph{slice}.
%
A minimal type error slice \citep{Haack2003-vc} includes all expressions that
are necessary for the error to manifest and no more.
%
We propose to use type error slices to communicate to the classifier
which expressions could potentially be blamed --- a change to an
expression outside of the minimal slice cannot possibly fix the type
error.
%
We will show that the type error slice is so important
(\autoref{sec:feature-utility}) that it is actually beneficial to
automatically discard expressions that are not part of the slice, rather
than letting the classifier learn to do so.
%
Our use of type error slices is conceptually related to the use of fault
localization in other program analysis and transformation techniques
(see \autoref{sec:related-work}).

\paragraph{\textbf{Challenge (2): What is the ``correct'' expression to blame?}}
%
%% As we saw in \autoref{fig:sumList},
%% type errors admit many different
%% fixes to different expressions,
%% each resulting in a different program.
%% %
%% How is the type checker to judge
%% which expressions are more likely
%% to be incorrect than others?

\paragraph{\textbf{Solution: Use the students' subsequent fixes}}
%
Rather than relying on expert users, we follow
\citet{Zhang2014-lv} \RJ{CHECK} and allow our students to
do the work of assigning blame themselves.
%
Software development is an iterative process and our students
eventually fix their own programs, perhaps after multiple ill-fated
attempts.

\paragraph{\textbf{Challenge (3): How can we intuitively present our feedback?}}
We now have the features and labels needed to train a classifier to
predict, for each expression in an ill-typed program, whether it should
be blamed.
%
But this is not yet particularly suitable as \emph{user feedback}.
%
A recent survey of developers by \citet{Kochhar2016-oc} found that
developers are unlikely to examine more than around five potentially
erroneous locations before falling back to manual debugging.
%
Thus, we should limit our predictions to a select few to be presented to
the user.

\paragraph{\textbf{Solution: Rank blame locations by ``confidence''}}
Luckily, many machine learning classifiers produce not only a predicted
label, but also a metric that can be interpreted as the classifier's
\emph{confidence} in its prediction.
%
Thus, we \emph{rank} each expression by the classifier's confidence that
it should be blamed, and present only the top-$k$ predictions to the
user (in practice $k=3$). This use of ranking to report
the results of a program analysis is popular in other problem domains
\citep[see, \eg][]{Kremenek2003-ck}; we focus
explicitly on the use of data-driven machine learning confidence as a
ranking source.

>>>>

The first step is to create a training
set of ill-typed programs, where each
erroneous sub-term is explicitly labeled.
%
Prior work has often enlisted
expert users to curate a set of
ill-typed programs and then
\emph{manually} determine the
correct fix~\citep[\eg][]{Lerner2007-dt,Loncaric2016-uk}.
%
This method is suitable for
\emph{evaluating} the quality
of a localization (or repair)
algorithm on a small number
(\eg, 10-100s) of programs.
%
However, in general it requires
a great deal of effort for the
expert to divine the original
programmers' intentions.
%
Consequently, is difficult to
scale the expert-labeling to
yield a dataset large enough
(\eg, 1000s of programs) needed
to facilitate machine learning.
%
More importantly, this approach
fails to capture the \emph{frequency}
with which errors occur in practice.

\mypara{Solution: Interaction Traces}
%
We solve both the scale and
frequency problems by instead
extracting blame-labeled data sets
from \emph{interaction traces}.
%
Software development is an iterative process.
Programmers, perhaps after a lengthy (and
sometimes frustrating) back-and-forth with
the type checker, eventually end up fixing
their own programs.
%
Thus, we implemented an instrumented
version of the \ocaml compiler that records
this conversation, \ie records the sequence
of programs submitted by each programmer and
whether or not it was deemed type-correct.
%
For each ill-typed program in
a particular programmer's trace,
we find the \emph{first subsequent}
program in the trace that type checks
and declare it to be the fixed version.
%
From this pair of an ill-typed program
and its fix, we can extract a \emph{diff}
of the abstract syntax trees, and then assign
the blame labels to the \emph{smallest}
sub-tree in the diff.

\mypara{Example}
For example, suppose our student
fixed the |sumList| program in
\autoref{fig:sumList} by replacing
|[]| with |0|, the diff would
include only the |[]| expression.
%
Thus we would determine that the
|[]| expression (and \emph{not} the
|+| or the recursive call |sumList t|)
is to blame.

@@

Next, we must find a way to translate
highly structured and variable sized
\emph{programs} into fixed size
$n$-dimensional numeric \emph{vectors}
that are needed for supervised
classification.
%
While the PL literature is full
of different kinds of ways to
represent programs, from raw
sequences of tokens, to
richly-structured
abstract syntax trees (AST) or
control-flow graphs, it is
unclear how to embed the
above into a vector space.
%
Furthermore, it is unclear whether
recent program representations that
are amenable to one learning task,
\eg code completion \cite{Devanbu:2012,Raychev:2014}
or decompilation \cite{Raychev2015-jg,Bielik2016-br}
are suitable for our problem of
assigning blame for type errors.

\mypara{Solution: Bags-of-Abstracted-Terms}
%
We present a new representation of programs
that draws inspiration from the theory of
abstract interpretation~\cite{CousotCousot77}.
%
Our representation as parameterized by a
set of \emph{feature abstraction} functions,
(abbreviated to feature abstractions)
$f_1, \ldots, f_n$, that map terms to a
numeric value (or just $\{0, 1\}$ to
encode a Boolean property.)
%
Given a set of feature abstractions, we
can represent a single program's AST as
a \emph{bag-of-abstracted-terms} (BOAT)
by:
%
(1)~Decomposing the AST (term) $t$ into
    a \emph{bag} of its constituent sub-trees
    (terms) $\{t_1,\ldots,t_m\}$, and then
%
(2)~Representing each sub-term $t_i$
    with the $n$-dimensional
    vector $[f_1(t_i),\ldots, f_n(t_i)]$.
%
Our choice of working with ASTs is natural
as that is the representation over which the
type-checker operates.

\mypara{Modeling Contexts}
%
Each expression occurs in some surrounding
\emph{context}, and we would like the
classifier to be able make decisions based
on the context as well.
%
The context is particularly important
for our task as each expression
imposes typing constraints on its
neighbors.
%
For example, \eg a |+| operator tells
the type checker that both \emph{children}
must have type |int| and that the \emph{parent}
must accept an |int|.
%
Similarly, if the student wrote
|h sumList t| \ie forgot the |+|,
we might wish to blame the application
rather than |h| because |h|
\emph{does not} have a function type.
%
The BOAT representation makes it
easy to incorporate contexts: we
simply \emph{concatenate} each
term's feature vector with the
\emph{contextual features} of
its parent and children.

%% \RJ{confusing -- delete}
%% We might expect that the recursive
%% call |sumList t| is unlikely to be
%% at fault because |sumList| has a
%% function type --- the user may have
%% called the wrong function or
%% supplied the wrong argument, but the
%% application itself is probably correct.

@@

Next, we must find a \emph{good}
set of features, that is, a set
of features that yields predictive
models. Our BOAT representation
lets us iteratively solve this
problem by just starting with
a simple set of features, and
then repeatedly adding more
and more to capture important
aspects needed to improve precision.
%
Our set feature abstractions
includes those that capture the
\emph{syntax}, \emph{types} and
\emph{context} of each expression.

\mypara{Syntax and Type Features}
%
We start by observing that
at the very least, the
classifier should be able
to distinguish between the
|[]| and |+| expressions
in \autoref{fig:sumList}
because they represent
different \emph{syntactic}
expression forms.
%
We model this by
introducing feature
abstractions of the form
is-nil, is-plus, \etc for
each of a fixed number of
primitive literal terms.
%
Modeling the syntactic class of an
expression gives the classifier a
basic notion of the relative
frequency of blame assignment
for the various program elements,
\ie perhaps |[]| is
\emph{empirically} more
likely to be blamed than |+|.
%
Similarly, we can model
the \emph{type} of each
sub-expression with features
of the form is-int, is-bool, \etc.
%
We will discuss handling
arbitrary, user-defined types
in~\autoref{sec:discussion}.

\mypara{Contextual Features: Error Slices}
%
Our contextual features include the
syntactic class of the neighboring
expressions, their inferred types
(when available).
%
However, in we have found that
the most important contextual
signal is whether or not the
expression occurs in a
\emph{minimal type error slice}~\citep{Haack2003-vc}
which includes the smallest
subset of all expressions
that are necessary for the
error to manifest.
%
We propose to use type error slices
to communicate to the classifier
which expressions could
\emph{potentially} be blamed --- a
change to an expression outside of
the minimal slice cannot possibly
fix the type error.
%
We empirically demonstrate that
the type error slice is so
important (\autoref{sec:feature-utility})
that it is actually beneficial to
automatically discard expressions
that are not part of the slice,
rather than letting the classifier
learn to do so.
%
Indeed, this domain-specific
insight is crucial for learning
classifiers that significantly
outperform the state-of-the-art.

\mypara{Example}
%
When \toolname is tasked with localizing
the error in the example program of \autoref{fig:sumList},
the |[]| and |+| sub-terms will each be given
their own feature vector, and we will ask the
classifier to predict for each \emph{independently}
whether it should be blamed.
%
\autoref{tab:sumList} lists some
of the sub-expressions of the example
from \autoref{fig:sumList}, and their
corresponding feature vectors.

% Our use of type error slices is
% conceptually related to the use
% of fault localization in other
% program analysis and transformation
% techniques (see \autoref{sec:related-work}).

@@

Finally, after we have trained a classifier
using the labeled data set, we need to use
it to help users localize type errors in
their programs.
%
The classifier tells us whether or not
a given sub-expression \emph{should be}
blamed (\ie has the blame label) but this
is not yet particularly suitable as
\emph{user feedback}.
%
A recent survey of developers by
\citet{Kochhar2016-oc} found that
developers are unlikely to examine
more than around five potentially
erroneous locations before falling
back to manual debugging.
%
Thus, we should limit our predictions
to a select few to be presented to
the user.

\mypara{Solution: Rank Locations by ``Confidence''}
%
Fortunately, many machine learning
classifiers produce not only a predicted
label, but also a metric that can be
interpreted as the classifier's
\emph{confidence} in its prediction.
%
Thus, we \emph{rank} each expression
by the classifier's confidence that
it should be blamed, and present only
the top-$k$ predictions to the
user (in practice $k=3$).
%
This use of ranking to report the
results of a program analysis is
popular in other problem domains
\citep[see, \eg][]{Kremenek2003-ck};
we focus explicitly on the use of
data-driven machine learning
confidence as a ranking source.
%
In \autoref{sec:evaluation} we show
that \toolname's ranking approach
yields a high-precision localizer:
when the top-3 locations are considered,
one of them is the real location
more than 90\% of the time.

%%% HOOK TO EVALUATION: Evaluation
